Simplified-spatial-availability-example

### 2022-05-26 Simplified-spatial-availability-example.Rmd
AP started reviewing the code; important change: we need to use the survey expansion factors!

We must manipulate the data before we can fit the impedance function:
```{r data-for-impedance}
#here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different imepedance function....
transit_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 9) %>%
  dplyr::select(TIEMPO_VIAJES)

car_motor_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>%
  dplyr::select(TIEMPO_VIAJES) #includes both motorcycles and cars

bike_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
  dplyr::select(TIEMPO_VIAJES)

walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>%
  dplyr::select(TIEMPO_VIAJES)
```

Then create the impedance functions:
```{r fitting-impedance-function}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
HW_transit_lnorm <- lnorm_

gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead")
HW_car_motor_gamma <- gamma_

lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
HW_bike_lnorm <- lnorm_

lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
HW_walk_lnorm <- lnorm_
```

Then we add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
```{r}
OD_v_HW <- OD_v_HW  %>% mutate(trips_transit = ifelse(MODO_PRIORITARIO <= 9, 1,0),
                    trips_car_motor = ifelse(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10, 1, 0),
                    trips_bike = ifelse(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20, 1, 0),
                    trips_walk = ifelse(MODO_PRIORITARIO == 24, 1, 0))

workers <- OD_v_HW %>%
  group_by(VORIZT1259) %>%
  summarize(workers_transit = sum(trips_transit),
            workers_car = sum(trips_car_motor),
            workers_bike = sum(trips_bike),
            workers_walk = sum(trips_walk),
            .groups = "drop") %>%
  transmute(ZT1259 = VORIZT1259,
            workers_transit,
            workers_car,
            workers_bike,
            workers_walk)

jobs <- OD_v_HW %>%
  group_by(VDESZT1259) %>%
  summarize(jobs_transit = sum(trips_transit),
            jobs_car = sum(trips_car_motor),
            jobs_bike = sum(trips_bike),
            jobs_walk = sum(trips_walk),
            .groups = "drop") %>%
  transmute(ZT1259 = VDESZT1259,
            jobs_transit,
            jobs_car,
            jobs_bike,
            jobs_walk)
```

Next join workers and jobs to the ZN simple features object:
```{r}
load("data/zn.RData")

zn <- zn %>%
  # Join workers
  left_join(workers,
            by = "ZT1259") %>%
  # Join jobs
  left_join(jobs,
            by = "ZT1259") %>%
  # Replace NAs with zeros
  mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
         workers_car = tidyr::replace_na(workers_car, 0),
         workers_bike = tidyr::replace_na(workers_bike, 0),
         workers_walk = tidyr::replace_na(workers_walk, 0),
         jobs_transit = tidyr::replace_na(jobs_transit, 0),
         jobs_car = tidyr::replace_na(jobs_car, 0),
         jobs_bike = tidyr::replace_na(jobs_bike, 0),
         jobs_walk = tidyr::replace_na(jobs_walk, 0))

>>>>

We must manipulate the data before we can fit the impedance function:
```{r data-for-impedance}
#here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different imepedance function....
transit_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 9) %>%
  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2)
  transmute(TIEMPO_VIAJES, 
            ELE_G_POND_ESC2,
            mode = "transit")

car_motor_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>%
  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2) #includes both motorcycles and cars
  transmute(TIEMPO_VIAJES, 
            ELE_G_POND_ESC2,
            mode = "car/motor")

bike_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2)
  transmute(TIEMPO_VIAJES, 
            ELE_G_POND_ESC2,
            mode = "bike")

walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>%
  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2)
  transmute(TIEMPO_VIAJES, 
            ELE_G_POND_ESC2,
            mode = "walk")
```

Then create the impedance functions:
```{r fitting-impedance-function}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,#%>% unlist() %>% as.numeric(),
                                "lnorm", 
                                method="mle",
                                #Notice the use of the weights,
                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
HW_transit_lnorm <- lnorm_

gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
                                "gamma", 
                                method="mle",
                                #Notice the use of the weights,
                                weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2), 
                                optim.method="Nelder-Mead")
HW_car_motor_gamma <- gamma_

lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
                                "lnorm", 
                                method="mle",
                                #Notice the use of the weights,
                                weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
HW_bike_lnorm <- lnorm_

lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
                                "lnorm", 
                                method="mle",
                                #Notice the use of the weights,
                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
HW_walk_lnorm <- lnorm_
```

To visualize the impedance functions we create a data frame and calculate the density of the distribution given the estimated parameters:
```{r}
x <- data.frame(t = seq(1, 300, 1))

x_transit <- x %>%
  mutate(f = dlnorm(t,
                    HW_transit_lnorm$estimate[1],
                    HW_transit_lnorm$estimate[2]),
         mode = "transit")

x_car_motor <- x %>%
  mutate(f = dgamma(t,
                    shape = HW_car_motor_gamma$estimate[1],
                    rate = HW_car_motor_gamma$estimate[2]),
         mode = "car/motor")

x_bike <- x %>%
  mutate(f = dlnorm(t,
                    HW_bike_lnorm$estimate[1],
                    HW_bike_lnorm$estimate[2]),
         mode = "bike")

x_walk <- x %>%
  mutate(f = dlnorm(t,
                    HW_walk_lnorm$estimate[1],
                    HW_walk_lnorm$estimate[2]),
         mode = "walk")
```

Plot:
```{r}
rbind(x_transit,
      x_car_motor,
      x_bike,
      x_walk) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
  ggplot(aes(x = t, y = f)) + 
  geom_line() + 
  facet_wrap(~ mode, nrow = 2)
```

What do the impedance curves tell us about the travel behavior by different modes? Check the histograms:
```{r}
rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
  ggplot(aes(x = TIEMPO_VIAJES, weights = ELE_G_POND_ESC2, after_stat(density))) + 
  geom_histogram(bins = 35) + 
  facet_wrap(~ mode, nrow = 2)
```

Then we add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
```{r}
# OD_v_HW <- OD_v_HW  %>% 
#   mutate(trips_transit = ifelse(MODO_PRIORITARIO <= 9, 1,0),
#          trips_car_motor = ifelse(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10, 1, 0),
#          trips_bike = ifelse(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20, 1, 0),
#          trips_walk = ifelse(MODO_PRIORITARIO == 24, 1, 0))

# Instead of creating various columns like above, create a single column with the mode
OD_v_HW <- OD_v_HW  %>% 
  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car_motor",
                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
                          MODO_PRIORITARIO == 24 ~ "walk"),
         mode = factor(mode))

# workers <- OD_v_HW %>%
#   group_by(VORIZT1259) %>%
#   summarize(workers_transit = sum(trips_transit),
#             workers_car = sum(trips_car_motor),
#             workers_bike = sum(trips_bike),
#             workers_walk = sum(trips_walk),
#             .groups = "drop") %>%
#   transmute(ZT1259 = VORIZT1259,
#             workers_transit,
#             workers_car,
#             workers_bike,
#             workers_walk)

# The workers are classified by mode
workers <- OD_v_HW %>%
  group_by(VORIZT1259,
           mode) %>%
  # The sum is of the weighted workers
  summarize(workers = sum(ELE_G_POND_ESC2),
            .groups = "drop") %>%
  rename(ZT1259 = VORIZT1259)

# jobs <- OD_v_HW %>%
#   group_by(VDESZT1259) %>%
#   summarize(jobs_transit = sum(trips_transit),
#             jobs_car = sum(trips_car_motor),
#             jobs_bike = sum(trips_bike),
#             jobs_walk = sum(trips_walk),
#             .groups = "drop") %>%
#   transmute(ZT1259 = VDESZT1259,
#             jobs_transit,
#             jobs_car,
#             jobs_bike,
#             jobs_walk)

# Do we need to classify the jobs by mode?
jobs <- OD_v_HW %>%
  group_by(VDESZT1259) %>%
  summarize(jobs = sum(ELE_G_POND_ESC2),
            .groups = "drop") %>%
  rename(ZT1259 = VDESZT1259)
```

NOW WE NEED: the travel times from the routing algorithm between origin-destination pairs. Anastasia did not use the routing, it may not be needed; instead she used the average travel time by mode for all trips

<!--
Next join workers and jobs to the ZN simple features object:
```{r}
load("data/zn.RData")

# zn <- zn %>%
#   # Join workers
#   left_join(workers,
#             by = "ZT1259") %>%
#   # Join jobs
#   left_join(jobs,
#             by = "ZT1259") %>%
#   # Replace NAs with zeros
#   mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
#          workers_car = tidyr::replace_na(workers_car, 0),
#          workers_bike = tidyr::replace_na(workers_bike, 0),
#          workers_walk = tidyr::replace_na(workers_walk, 0),
#          jobs_transit = tidyr::replace_na(jobs_transit, 0),
#          jobs_car = tidyr::replace_na(jobs_car, 0),
#          jobs_bike = tidyr::replace_na(jobs_bike, 0),
#          jobs_walk = tidyr::replace_na(jobs_walk, 0))

workers_sf <- workers %>%
  # Join workers
  left_join(zn,
            by = "ZT1259") %>%
  st_as_sf()

jobs_sf <- jobs %>%
  # Join jobs
  left_join(zn,
            by = "ZT1259") %>%
  st_as_sf()
